---
title: "Week 44 Lecture 1"
author: "562770bf"
date: "2024-10-28"
output: html_document
---
This is the first Rmd file for the Data Science and HR analytics course, trying out experiments on the code for lecture 1 in week 44.

So let's get started. 
```{r}
CEO_Diary <- read.csv("~/Dropbox/Mijn Mac (Macintosh.local)/Downloads/2017765data/behaviors/python_index-construction/baseline_results/survey_response_data.csv")
```

This command reads the csv file and calls the data in the table CEO Diary. The view command (not included as that does not work in a markdown file) just shows a visual overview of the data in a table. 

```{r}
CEO_Diary[1:15,c(1:5,37, 39, 40)] 
```
This command shows a table of the previously inserted data, the first term shows the number of rows, 1 through 15 and the next term specifies the columns shown (1 through 5, 37, 39 and 40). Hence you can easily change the specifications to show a different number of rows and different variables as columns.
```{r}
CEO_Diary[1:7,c(6, 15, 21, 27, 35)] 
apply(CEO_Diary,2,class)
```
The apply class command shows what sort of type the variables are, most of which are defined as string variables. The number 2 specifies that it will define the type for the columns. If you choose 1 instead of 2, it will do so for the rows. If you choose mean/max/min instead of class, it will define those values instead, which does not work here as most variables are all string variables. 
```{r}
nrow(CEO_Diary)
summary(CEO_Diary[1:5])
summary(CEO_Diary[5:10])
summary(CEO_Diary[37])
```
The nrow command shows the number of rows in the data. The summary command shows a summary of the specified columns/variable range. 
```{r}
  png(file="~/Dropbox/Mijn Mac (Macintosh.local)/Downloads/CEO_Types.png", width=800, height=300)
  par(mar=c(9, 3 ,1,1))
  barplot(prop.table(table(CEO_Diary$type)), las=2)
  dev.off()
```
The code above saves a PNG file showing the relative frequency of each CEO type in CEO_Diary$type. The barplot function plots proportions, and the las=2 argument rotates axis labels for readability.

```{r}
  table(CEO_Diary$type)
  prop.table(table(CEO_Diary$type))
  barplot(prop.table(table(CEO_Diary$type)))
  barplot(prop.table(table(CEO_Diary$type)), las=2)
```
The table command creates a frequency table for the type variable in the CEO_Diary dataset.The result shows the count of each unique value in the type column, helping you understand the distribution of CEO activity types.
The prop.table converts the frequency counts into proportions, showing the relative frequency of each type category as a fraction of the total count (i.e., each category’s share in decimal form).
The third line of code creates a bar plot of the proportions calculated in the previous step.Each bar represents a different type of CEO activity, with the bar height showing the proportion of that activity type relative to the total.
The final line of code repeats the bar plot, but with las=2, which rotates the x-axis labels to be perpendicular to the axis. This rotation can make labels more readable if there are many categories or if the labels are lengthy.
```{r}
CEO_Diary <- read.csv("~/Dropbox/Mijn Mac (Macintosh.local)/Downloads/2017765data/behaviors/python_index-construction/baseline_results/survey_response_data.csv")
fit <- glm(strategy ~ consultants + politicians, data=CEO_Diary); summary(fit)
```

This command fits a regression model, depending on the data type of strategy) where:
strategy is the dependent variable or outcome.Consultants and politicians are the independent variables.
'Data=CEO_Diary' specifies that the data for this model comes from the CEO_Diary dataset.
The glm term is a function for fitting generalized linear models, which includes a variety of model types (e.g., logistic regression for binary outcomes or linear regression for continuous outcomes).

```{r}
fit2 <- glm(strategy ~ govoff + unions, data=CEO_Diary); summary(fit2)
```
We get different estimates, both coefficients are now negative, but still small. However, hard to say anything useful here as there likely is no causal interpretation (significance of coefficients does not say that much about truth I think).
